A conditional permutation-based approach to test confounder effect and center-bias in machine learning models

### 2021-08-13 manuscript.tex
patched
ready for inner review

\begin{abstract} % 150 words
The lack of rigorous non-parametric statistical tests of confounder-effects significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal.
Simulations show that the tests are applicable in a wide variety of cases with a remarkable statistical power.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset identified significantly biased predictions. Some of these biases were previously unreported and found to be hard to mitigate with state-of-the-art confound mitigation approaches.
The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
\end{abstract}

@@

Predictive modelling and supervised learning methods have recently became increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}. When evaluate the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) i snot the only consideration. Generalizability across contexts and populations and biomedical validity (i.e. whether the model is driven by biomedical relevant signal) are also highly important requirements for candidate biomarkers \citep{woo2017building}.

Variables that can induce spurious associations between the predictor variables (features) and the prediction target - often referred to as confounders - can be detrimental to the model's biomedical validity and generalizability. Many types of confounders can be distinguished.
First, measurement artifacts can obviously be considered as confounders. For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with movement kinematics or motor function, like previously demonstrated, among others, in case of e.g Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated}er or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}. Similarly, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling. For instance, resting-state functional connectivity is known to be strongly predictive to  demographic age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}. As fluid intelligence is known to decline with aging, \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by simply picking up on age-related variance \citep{lohmann2021predicting, dubois2018distributed}. 
Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance}. In this case, predictive models can utilize center/batch-effects to explain (random or true) center/batch-differences in the predictive target. Such models will display worse-than-expected performance when applied on data from new centers/batches.

Recently, various data harmonization approaches have been suggested to mitigate confounder-bias, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}. However, it is often unclear which variables should be considered as confounders during harmonization and considering many variables as confounder holds risks of eliminating signal-of-interest \citep{wachinger2021detect}.

Powerful statistical tests to quantify the confounding effect of certain variables in predictive models could largely foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches. Such methods, however, have to tackle the fact that the outputs of predictive models often follow non-normal distributions (due to e.g. regularization) \citep{garcia2009study, kristensen2017whole}.

Recently three dedicated techniques \citep{chaibub2019permutation, ferrari2020measuring, wachinger2021detect} have been proposed for quantifying confounder bias in predictive modelling.

The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via restricted permutations. It's p-values, however, might be invalid in certain circumstances (e.g. biased classification problems) and, as shown by \cite{ferrari2020measuring}, it does not guarantee a proper type I error control.
The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, by design, does not provide any quantification of statistical significance.
Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the the predictors are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both. It's computation assumes 'faithfulness' (if two variables are independent, there is no direct influence between the two in the underlying graph) a strong condition that might not be fulfilled in many real applications (e.g. functional MRI is an \emph{indirect} measure of neural activation, rendering the alternative hypothesis of this test true in absence of any real confounder).

@@

Next, I utilize the permutation-based conditional independence framework of \cite{berrett2020conditional}, to construct two novel tests for evaluating the confounder-bias in predictive models.
The \emph{full confounder test} probes whether the model's explained variance can be attributed exclusively to the confounder
The \emph{partial confounder test} investigates whether the model utilizes any confounder-variance, above what is already present in the target variable.
Both tests require only the target variable, the model predictions and the putative confounder as input, that is, they can be performed without needing to re-fit the model.
The tests set no assumption about the distribution of the model predictions.
The validity of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normal model output).
The proposed tests are then applied on functional brain connectivity data to provide evidence of motion-bias in a model trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} and to evaluate the age dependence of fluid intelligence predictions on data from the Human Connectome Project.

@@

In a predictive modelling setting, let $\y$ denote the target variable, $\yhat$ denote model output, i.e. the predictions for $\y$ and let $\c$ denote a variable which is considered as a confounder (see Fig. \ref{fig:schematic}B). Let us assume, furthermore, that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i)$ for $i=1, \dots , n$ so that $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.

@@

Instead of focusing on the unconditional independence between the confoudner and the predictions, we shall consider the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, according to \cite{dawid1979conditional}, implies that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$. The test with this null hypothesis will be referred to as the \emph{partial confounder test}.

Alternatively, one might be also interested in testing $\yhat \independent \c | \y$. We will refer to the corresponding test as the \emph{full confounder test}.

@@

Even if constructing totally non-parametric confoudner tests was proven to be impossible, as shown by \cite{berrett2020conditional} and \citep{candes2016panning}, valid and powerful conditional independence tests can be constructed with inputting information about the distribution of only two (out of the three) variables.
Specifically, the conditional permutation test (CPT) of Berret and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable, thereby it incorporating available distributional information into permutation-based inference.

@@

Within this setting, conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(\y, \yhat, \c)$ triplet, as listed in Table \ref{tab:conditional-independence-cases}. Testing the null hypothesis $\y \independent \yhat | \c$ (option 1, full confoudner testing) investigates whether the model predictions are likely explainable solely with the confounder, i.e. whether the model is exclusively confounder-driven. Testing $\y \independent \c | \yhat$ (option 2) addresses the question whether the model captures all the variance in $c$ when predicting $y$. Testing the null hypothesis $\yhat \independent \c | \y$ (option 3, partial confounder testing) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e. whether there is any confounder bias in the model.

@@

Option 3, i.e. partial confounder testing is typically of interest when testing confounder bias of predictive models. Option 1, i.e. full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction. Option 2 seems less appealing for model diagnostics and importantly, in this case the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.

@@

To ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:

@@

are all identically distributed and exchangeable, and so are the

@@

The p-value under the null hypothesis is then obtained as
$$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$

For a verification of the valid type I error control of this approach refer to Theorem 1 in \citep{berrett2020conditional}.

@@

\begin{equation}
\frac{ q(c_j | y_i) q(c_i | y_j)}{q(c_i | y_i) q(c_j | y_j) }
=
log \mathcal{L}(c_j | y_i) + log \mathcal{L}(c_i | y_j) - log \mathcal{L}(c_i | y_i) - log \mathcal{L}(c_j | y_j) 
\label{eq:accept-odds}
\end{equation}

@@

Obtaining a relatively accurate, independent estimate of $D(\cdot|\y)$ (of any shape) for CPT inference is important. Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix of the conditional distribution $D(\cdot|Z)$, which might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
Since in the proposed predictive modelling, normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.

@@

with the assumption that the residual error follows a Gaussian distribution. If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma$ is the standard deviation of the residual $\boldsymbol{e}$, then the conditional distribution of interest is given as $ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$ and the log-likelihood, that is to be used in Eq. \ref{eq:accept-odds}, can be computed simply as the log of the corresponding probability density function:

$$ log\mathcal{L}(c_i|y_j) = - \frac{1}{2} \Big(\frac{c_i-\mu_j}{\sigma}\Big)^2 - ln(2 \pi \sigma)   $$

In the case of categorical target, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$.

Notably, from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, the proposed Gaussian likelihood estimation is possible for options 1 and 3, but not option 2, as the variable on which the independence is conditional must be always the predictor variable in Eq. \ref{eq:linreg}.
For options 1 and 3, the proposed simple Gaussian conditional likelihood estimation allows for deploying CPT without a large, independent sample to estimate the conditional distribution.

While, in theory, any predefined test statistic $T$ can be used, for the proposed Gaussian likelihood estimation, the python package \emph{mlconfound} implements only the two-sided variant of the tests (as on-sided testing is out-of-interest in the confoudner problem) and uses the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification) as a test statistic:

@@

for any $\pi \in S_n$.

@@

Here I extend their results by performing simulations for estimating the power of the test in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\yhat$ and in the case of varying levels of covariance between $\y$ and $\c$).
Moreover, I perform a thorough investigation of the robustness of the test against the violation of normality (for both $(\y, \c)$ and $\yhat$).

@@

\subsubsection*{Gaussian Simulations}
The target variable $\y$ and the confounder variable $\c$ were drawn randomly from a multivariate normal distribution:

@@

where $\mu=(0, 0)$ and the covariance matrix is given as:
$\Sigma_\sigma = \big(\begin{smallmatrix}
  1 & \sigma\\
  \sigma & 1
\end{smallmatrix}\big)$.

Next, the simulated predicted values are constructed:

$$ \yhat = w_y \y + w_c \c + w_e \boldsymbol{e}$$

@@

For more straightforward interpretation and visualization of the simulation results, the 'confound-to-target' and the 'signal-in-prediction' parameters were defined as

@@

Next to the normally distributed case, four \emph{sinh-arcsinh} transformed cases were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$

@@

\subsection{Validation on functional brain connectivity data}

The usefulness of the proposed confounder tests is demonstrated by applying them for predictive classification and regression models based on functional brain connectivity data, with different confound-mitigation approaches.

@@

The proposed tests were applied to provide evidence of motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464. Preprocessed regional timeseries data was obtained as shared with the paper \citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project \citep{craddock2013neuro}.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.40\paperwidth]{fig/sim_non-norm.eps}
  \caption{\textbf{Robustness of conditional permutation based confound testing to the violation of normality assumptions for $Y$ and $C$.} \\
  Simulations included variables with five different degrees of non-normality (A), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation, from normal (left) to severely skewed and heavy tailed (right). Fisher's kurtosis and skewness is given for each at the bottom of panel A. Panel (B) depicts false positive rates in the simulations under H0, separately for different signal-in-confounder ratios (here identical to $w_{Y}$).
  Vertical axis shows type I error rate. The nominal type I error rate of $alpha=0.05$ is shown with a vertical dashed line. Panel (C) shows how violation of the normality assumption affects power. Boxplots show the mean power over simulation cases, grouped by confound-to-target ratio and sample size.}
  \label{fig:sim-non-normal}
\end{figure}

@@

Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected, completely non-parametric regarding the distribution of the predictions $\yhat$). The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal}, involved various \emph{sinh-arcsinh} distributions, form normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to kurtosis and skewness values as high as 5 and 2 respectively. Furthermore, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).

@@

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
The partial confounder test revealed that, the model was significantly biased both by age group and acquisition batch (both $p<0.0001$) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.084$ and $0.041$, respectively) associated to larger predicted values.

After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all). Both feature regression and COMBAT decreased the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=0.75$ and $0.76$ when applied to remove the effect of acquisition and age, respectively).

@@

Importantly, the simulations strongly suggest that both tests are very robust to the violation of the normality assumption for the conditional distribution between the target and the confounder. Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extreme violations of normality (see the purple distribution on Fig. \ref{fig:sim-non-normal}). Due to the conservatives of the tests, if predictive model is very accurate, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates.

Non-normality has a negative effect on the power of the both tests, however, the decrease in power is in many simulation cases negligible and gets only relevant if the sample size and/or the confounder-effect is low.

Given the power and the robustness of the proposed tests, they might become a useful tool in diagnosing predictive models in a wide spectrum of applications.

>>>>

\begin{abstract} % 150 words
The lack of rigorous non-parametric statistical tests of confounder-bias significantly hampers the development of robust, valid and generalizable predictive models in many fields of research.
Here I propose the \emph{partial} and \emph{full confounder tests}, that build on a recently established theoretical framework for permutation-based conditional independence testing and probe the null hypotheses of \emph{unbiased} and \emph{fully biased model}, respectively.
The proposed tests set no assumptions about the distribution of the predictive model output that is often non-normal and display a remarkable robustness and a high statistical power in simulations.
Applying the proposed tests on models trained on functional brain connectivity data from the Human Connectome Project and the Autism Brain Imaging Data Exchange dataset provided statistical evidence for several models being significantly confounded. Some of these confounder-biases were previously unreported and found to be hard to fully mitigate with state-of-the-art confound mitigation approaches.
The package \emph{mlconfound}\footnote{\href{https://mlconfound.readthedocs.io}{https://mlconfound.readthedocs.io}} can aid the assessment and improvement of the generalizability and neurobiological validity of predictive models and, thereby, foster the development of clinically useful machine learning biomarkers.
\end{abstract}

@@

Predictive modelling and supervised learning methods have recently became increasingly important in biomedical research and hold promise for delivering biomarkers that substantially impact clinical practice and public health \citep{kent2018personalized}. When evaluating the usefulness and applicability of such markers, predictive performance (i.e. prognostic/diagnostic value) is far from being the only important consideration. Generalizability across contexts and populations and biomedical validity (i.e. whether the model is driven by biomedically relevant signal) are also crucial requirements for candidate biomarkers \citep{woo2017building}.

Variables that can induce spurious associations between the predictor variables (features) and the prediction target - often referred to as confounders - can be detrimental to the model's biomedical validity and generalizability. Many types of confounders can be distinguished.
First, measurement artifacts can obviously be considered as confounders. For instance, in-scanner head motion artifacts in magnetic resonance imaging (MRI)-based predictive models can be especially problematic, if the prediction target is also associated with movement kinematics or motor function, like previously demonstrated, among others, in case of Alzheimer's disease \citep{rao2017predictive} attention deficit hyperactivity disorder (ADHD) \citep{couvy2016head, eloyan2012automated} or Autism Spectrum Disorder (ASD) \citep{spisak2014voxel, spisak2019optimal, gotts2013perils}. Similarly, eye-blink artifacts can result in confounder-bias in EEG-based markers of ASD  \citep{eldridge2014robust}.
Second, depending on the research question, demographic and psychometric variables can also be considered as confounders during predictive modelling. For instance, resting-state functional connectivity is known to be strongly predictive to  demographic age \citep{wang2012decoding, dukart2011age} and moderately to fluid intelligence \citep{he2020deep, cole2012global}. As fluid intelligence is known to decline with aging, \citep{kievit2018neural}, models trained to predict intelligence can provide reasonable performance by picking up on solely age-related variance \citep{lohmann2021predicting, dubois2018distributed}. 
Third, batch effects or, in multi-center studies, center-effects can also cause significant bias in predictive models \citep{leek2010tackling, da2020performance} as the models can utilize such effects to explain (random or true) center/batch-differences in the predictive target. Such models will display worse-than-expected performance when applied on data from new centers/batches.

Various data harmonization approaches have been proposed to mitigate confounder-bias, ranging from sample matching \citep{rao2017predictive} and regression-based techniques \citep{rao2017predictive, dukart2011age, spisak2014voxel, abdulkadir2014reduction} to the popular non-parametric empirical Bayes-based method of \cite{johnson2007adjusting} and sophisticated deep-learning oriented solutions \citep{zhao2020training, hognon2019standardization}. However, it is often unclear which variables should be considered as confounders during harmonization and such approaches hold risks of eliminating signal-of-interest \citep{wachinger2021detect}.

Powerful and robust statistical tests to quantify the confounding effect of certain variables in predictive models could largely foster both the the identification of confounders to correct for and the assessment of the effectiveness of various confound-mitigation and data harmonization approaches. Such methods, however, have to tackle the fact that the outputs of predictive models often follow non-normal distributions (due to e.g. regularization) \citep{garcia2009study, kristensen2017whole}.

Recently, three dedicated techniques \citep{chaibub2019permutation, ferrari2020measuring, wachinger2021detect} have been proposed for quantifying confounder bias in predictive modelling.

The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via "restricted permutations" that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature) However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances. Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control with biased classification problems.
The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, by design, does not provide any quantification of statistical significance.
Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the the predictors are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both. It's computation assumes 'faithfulness' (if two variables are independent, there is no direct influence between the two in the underlying graph) a strong condition that might not be fulfilled in many real applications (e.g. functional MRI is an \emph{indirect} measure of neural activation, rendering the alternative hypothesis of this test true in absence of any real confounder).

@@

Next, I utilize the permutation-based conditional independence framework of \cite{berrett2020conditional}, to construct two novel tests for evaluating the confounder-bias in predictive models.
The \emph{full confounder test} probes whether the model's predictive performance can be attributed exclusively to the confounder.
The \emph{partial confounder test} investigates whether the model utilizes any confounder-variance, when controlled for the target variable.
Both tests require only the target variable, the model predictions and the putative confounder as input, thus, they can be performed with a negligible extra computational cost (re-fitting the model is not required).
The tests set no assumption about the distribution of the model predictions.
The robustness of the type I error control and the statistical power of both tests are evaluated with simulated data (assuming non-normality).
The proposed tests are then applied on functional brain connectivity data to provide evidence of motion- and center-bias in a model trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} and to evaluate the age and acquisition-batch dependence of fluid intelligence predictions on data from the Human Connectome Project.

@@

In a predictive modelling setting, let $\y$ denote the target variable, $\yhat$ denote model output, i.e. the predictions for $\y$ and let $\c$ denote a variable which is considered as a confounder (see Fig. \ref{fig:schematic}B). Let us assume, furthermore, that $\y$, $\yhat$ and $\c$ consist of independent and identically distributed (IID) data points $(y_i, \hat{y}_i, c_i) \in Y \times \\hat{Y} \times C$ for $i=1, \dots , n$ so that $\y=(y_1, \dots ,y_n)$, $\yhat=(\hat{y}_1, \dots, \hat{y}_n)$ and $\c=(c_1, \dots, c_n)$.

@@

Instead of focusing on the unconditional independence between the confounder and the predictions, we shall consider the \emph{conditionally independence} between $\yhat$ and $\c$ given $\y$ (written as $\yhat \independent \c | \y$) which, by definition  \citep{dawid1979conditional}, means that $\mathbb{P}(\yhat, \c|\y) = \mathbb{P}(\yhat|\y)\mathbb{P}(\c|\y)$. The statistical test with this null hypothesis will be referred to as the \emph{partial confounder test}.

Alternatively, one might be also interested in testing $\yhat \independent \y | \c$. We will refer to the corresponding test as the \emph{full confounder test}.

@@

Even if constructing fully non-parametric confounder tests was proven to be impossible, as demonstrated by \citep{candes2016panning} and \cite{berrett2020conditional}, valid and powerful conditional independence tests can be constructed with inputting distributional information about only two (out of the three) variables.
Specifically, the conditional permutation test (CPT) of Berret and colleagues samples from a non-uniform distribution over the set of possible permutations $\pi$ of one of the variables, based on its conditional distribution of the other variable. Thereby, it incorporates the information available on the conditional distribution of interest into the permutation-based inference in a statistically valid manner.

@@

Within this setting, conditional independence testing and, specifically, the framework of conditional permutation testing allows investigating three different null hypotheses corresponding to the $(\y, \yhat, \c)$ triplet. As listed in Table \ref{tab:conditional-independence-cases}, testing the null hypothesis $\y \independent \yhat | \c$ (option 1, full confounder testing) investigates whether the model predictions are likely explainable solely with the confounder, i.e. whether the model is exclusively confounder-driven. Testing $\y \independent \c | \yhat$ (option 2) addresses the question whether the model captures all the variance in $c$ when predicting $y$. Testing the null hypothesis $\yhat \independent \c | \y$ (option 3, partial confounder testing) examines, whether the dependence of the model output on the confounder can likely be explained by the confounder's dependence on the target variable, i.e. whether there is any confounder bias in the model.

@@

Option 3, i.e. partial confounder testing is typically of interest when testing confounder bias of predictive models. Option 1, i.e. full confounder testing, might be also useful in diagnostics of predictive models, especially in the exploratory phase of model construction. Option 2 seems less appealing for model diagnostics and importantly, in this case the proposed variety of the CPT framework does not allow constructing a test which is non-parametric on $\yhat$.

@@

As shown in \citep{berrett2020conditional}, to ensure that Eq. \ref{eq:cond-copy} holds, the $\c_{\boldsymbol{\pi}^{(j)}}$ copies must be drawn so that:

@@

are all identically distributed and so are the

@@

As long as the numerator of Eq. \ref{eq-pperm} is non-zero for all $c_\pi \in C$ and $y \in Y$, the conditional permutations constitute an algebraic group, thus, as shown by \citep{hemerik2018exact}, an unbiased estimate of the p-value under the null can be obtained as:
$$ p= \frac{\sum_{j=1}^m \mathbb{1} \{T(\y, \yhat, \c^{(j)}) \geq T(\y, \yhat, \c) \}  }{m}$$

While the group property of the conditioned permutations provides a straightforward proof for the validity of the approach, for an alternative verification see the proof of Theorem 1 in \citep{berrett2020conditional}.

@@

\begin{equation}
\frac{ q(c_j | y_i) q(c_i | y_j)}{q(c_i | y_i) q(c_j | y_j) }
=
\ell(c_j | y_i) + \ell(c_i | y_j) - \ell(c_i | y_i) - \ell(c_j | y_j) 
\label{eq:accept-odds}
\end{equation}

where $\ell$ denotes the log-likelihood.

@@

Obtaining a relatively accurate, independent estimate of $D(\cdot|\y)$ (of any shape) for CPT inference is important. Berrett and colleagues recommend to use a large independent sample to obtain the log-likelihood matrix that represents the conditional distribution $D(\cdot|Z)$. This approach might be often viable in semi-supervised learning settings where unlabelled data $(X, Z)$ are easier to obtain than labelled data $(X, Y, Z)$.
Since in the proposed predictive modelling setting, normality might be often a reasonable assumption for $\y$ and $\c$ (but not for $\yhat$), it is possible to fall back to Gaussian assumptions and thereby eliminate the need for a large, independent sample to estimate $D(\cdot|\y)$.

@@

with the assumption that the residual error follows a Gaussian distribution. If we write $\boldsymbol{\mu} = \alpha + \beta \y$ and $\sigma$ denotes the standard deviation of the residual $\boldsymbol{e}$, then the conditional distribution of interest can also be assumed to be normal with the parameters:

$$ (\c|\y=y_i) \sim \mathcal{N}\{\mu_i, \sigma^2\}$$

and the log-likelihood, that is to be used in Eq. \ref{eq:accept-odds}, can be computed simply as the log of the corresponding probability density function:

$$ \ell(c_i|y_j) = - \frac{1}{2} \Big(\frac{c_i-\mu_j}{\sigma}\Big)^2 - ln(2 \pi \sigma)   $$

In the case of categorical $c$, a multinomial logistic regression model can be used to obtain $D(\cdot|\y)$, with the extra assumption of non-complete separation if $y$ is also categorical (for an invertible Hessian, see e.g. \citep{bohning1992multinomial}).

Importantly, both the linear and the multinomial logistic regression approaches guarantee that the numerator of Eq. \ref{eq-pperm} is always greater than zero and the group property for the permutations holds.

Note that from the three options for conditional independence-based null hypotheses enumerated in Table \ref{tab:conditional-independence-cases}, for option 2, the proposed approach can not provide a test that is assumption-free about $yhat$, as the variable on which the independence is conditional must be always the predictor variable in Eq. \ref{eq:linreg}. However, as discussed above, this option is of low practical relevance, anyway.
Pleasingly,  the proposed Gaussian regression-based conditional likelihood estimation ensures that no assumptions on $\yhat$ have to be made for the practically relevant options 1 and 3, i.e. for the full and partial confounder tests.

In theory, any predefined test statistic $T$ can be used with the proposed approach. The python package \emph{mlconfound}, implementing the proposed full and partial confoudner tests, utilizes the coefficient of determination ($R^2$ or pseudo $R^2$ in case of categorical confounder or classification \citep{starkweather2011multinomial}) as a test statistic:

@@

which allows interpretable, two tailed inference about confoudner-bias in predictive modelling.

@@

While their findings are largely applicable to the proposed tests, here I extend those results by performing simulations in a form that is accessible for power calculations in predictive modelling (considering various weights of the confounder and the target signals in $\yhat$ and in the case of varying levels of covariance between $\y$ and $\c$).
Moreover, I perform a thorough investigation of the robustness of the test against the violation of normality (for both $(\y, \c)$ and $\yhat$).

@@

\subsubsection*{Gaussian Simulations}
As  a first step, the target variable $\y$ and the confounder variable $\c$ are drawn randomly from a multivariate normal distribution:

@@

where $\mu_0=(0, 0)$ and the covariance matrix is given as:
$\Sigma_\sigma = \big(\begin{smallmatrix}
  1 & \sigma\\
  \sigma & 1
\end{smallmatrix}\big)$ with $\sigma$ (i.e. the covariance between $\y$ and $\c$) being a free parameter of the simulation.

Next, the simulated predicted values are constructed:
 with an additive model:
$$ \yhat = w_y \y + w_c \c + w_e \boldsymbol{e}$$

@@

For more straightforward interpretation and visualization of the simulation results, the 'confound-to-target' and the 'signal-in-prediction' parameters are defined as

@@

Next to the above described normally distributed simulation scenario, four analogous, but \emph{sinh-arcsinh} transformed scenarios were also investigated, with $\delta \in \{1, 1.05, 1.5, 5\}$ $\epsilon \in \{1, 3, 5, 10\}$

@@

\subsection{Application on functional brain connectivity data}

The usefulness of the proposed confounder tests is demonstrated by applying them for predictive classification and regression models based on functional brain connectivity data, processed with different confound-mitigation approaches.

@@

The proposed tests were applied to provide evidence of center- and  motion-bias in diagnostic predictive models of autism spectrum disorder (ASD), trained on the Autism Brain Imaging Data Exchange (ABIDE) dataset \citep{di2014autism} involving 866 participants; ASD: 402, healthy control (HC): 464. Preprocessed regional timeseries data was obtained as shared with the paper \citep{dadi2019benchmarking} which was based on preprocessed image data provided by the Preprocessed Connectome Project \citep{craddock2013neuro}.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.40\paperwidth]{fig/sim_non-norm.eps}
  \caption{\textbf{Robustness of conditional permutation based confound testing to the violation of normality assumptions for $\y$ and $\c$.} \\
  Simulations included variables with five different degrees of non-normality (A), as introduced with various $\delta$ and $\epsilon$ values of the \emph{sinh-arcsinh} transformation, from normal (left) to severely skewed and heavy tailed (right). Fisher's kurtosis and skewness is given for each at the bottom of panel A. Panel (B) depicts false positive rates in the simulations under H0, separately for different signal-in-confounder ratios (here identical to $w_{y}$).
  Vertical axis shows type I error rate. The nominal type I error rate of $alpha=0.05$ is shown with a vertical dashed line. Panel (C) shows how violation of the normality assumption affects power. Boxplots show the mean power over simulation cases, grouped by confound-to-target ratio and sample size.}
  \label{fig:sim-non-normal}
\end{figure}

@@

Simulations with non-normal data showed that the proposed methods are remarkably robust to the violation of normality assumptions for the target variable $\y$ and the confounder $\c$ and, as expected, completely non-parametric regarding the distribution of the predictions $\yhat$). The six simulation cases of normality violation, shown on Figure \ref{fig:sim-non-normal}, involved various \emph{sinh-arcsinh} distributions, form normal (Fisher's kurtosis: 0, skewness) to severely skewed and heavy tailed (kurtosis: 42, skewness: 6).
The results show that the violation of normality - as expected - increases the type I error rate, but the increase - depending on the application - might not be severely problematic up to kurtosis and skewness values as high as 5 and 2 respectively. Furthermore, in terms of type I error rate, normality violation becomes less problematic in case of high predictive performance (high association between the predicted and observed values).

@@

Both acquisition batch and age group were statistically significantly associated with fluid intelligence ($R^2=0.032$ and $0.011$ and $p<0.001$ and $=0.001$, respectively, see also Table \ref{tab:unconditional-pvals}). The model trained on the raw (unadjusted) connectivity features predicted fluid intelligence with a medium effect size ($R^2=0.095$, $p<0.001$).
The partial confounder test revealed that the model was significantly biased both by age group and acquisition batch (both $p<0.0001$, first column of Fig. \ref{fig:hcp}) with later phases of the acquisition and lower age being weakly-moderately ($R^2=0.084$ and $0.041$, respectively) associated to larger predicted values.

After applying confound mitigation approaches (feature regression or COMBAT) the partial confound test did not provide evidence of confounder bias anymore ($p > 0.05$ for all, second and third columns of Fig. \ref{fig:hcp}). Both feature regression and COMBAT decreased the predictive performance in the case of both confounders, with COMBAT providing better performances ($R^2=0.75$ and $0.76$ when applied to remove the effect of acquisition and age, respectively).

@@

Next to their statistical power, the practical applicability of the proposed tests is also highly affected by their robustness of the normality assumptions on the conditional distribution of $\y$ and $\c$. Importantly, the simulations strongly suggest that both tests are strikingly robust to the violation of this assumption. Type I error is well controlled even in case of highly skewed and heavy tailed distributions (kurtosis > 5, skewness > 2) and gets significantly inflated only with extreme violations of normality (see the purple distribution on Fig. \ref{fig:sim-non-normal}). Due to the conservatives of the tests, if the predictive model is very accurate, even this extremely skewed and heavy tailed distribution resulted in lower than the nominal type I error rates.

Non-normality has a negative effect on the power of both tests, however, the decrease in power is in most cases small or even negligible and gets only more pronounced if the sample size and/or the confounder-effect is low.
